Assessing exposure to Atlantic Basin tropical storms in United States counties

### 2016-06-16 writing/DraftExposurePaper/DraftExposurePaper.tex
Close #36

  \title{Assessing exposure to Atlantic Basin tropical storms in United States
counties}
    \author[Colorado State University]{Joshua Ferreri}
   \ead{alice@example.com}

@@

@@

@@

@@

@@

It can be very difficult to reliabily measure rain during extreme rain
events, including tropical storms. For example, a heavy rain can wash
away {[}?{]} rain monitors {[}?{]}. It can also be very hard to measure
rain during heavy wind, as the rain does not fall straight into the
monitor {[}?{]}.

Some of the other possible sources for estimating rain during tropical
storms include \ldots{}

@@

Rainfall estimates are likely underestimates for a few reasons. First,
they are based on averaging hourly measurements to a daily mean
estimate. This averaging would smooth over shorter periods of very
extreme rainfall. Further, this data is averaged over multiple grid
points within each county and so would not fully reflect very extreme
local precipitation (although this might be less of a concern for
classifying exposure to a large-scale storm system, like a tropical
storm, compared to more fine-scale storms). Finally, this NLDAS data
provides a re-analysis that incorporates measured rainfall, using
models, etc., to incorporate that observed data into a spatially and
temporally continuous dataset of rainfall. However, during extreme
storms, the problems with measuring rainfall using {[}rain monitors{]}
would propogate into the NLDAS data, so although NLDAS would prevent
missing values during the storm if monitors are not able to provide
data, if monitors are out, rainfall estimates from NLDAS will be based
more on models than on observations.

Exposure classification based on rainfall has some advantages. For
example, it allows the identification of exposed counties that are
inland, rather than coastal, but that were affected by heavy rainfall
during the storm. Often, storm-related deaths are associated with inland
flooding. Storms can produce a lot of rain especially in certain
topographies, like near mountains, so counties that are well inland can
sometimes experience more extreme rain that other counties at similar
distances from the storm's track.

In comparison to the storm track, the areas of extreme rain and extreme
wind can vary. For example, storms undergoing an extratropical
transition can bring heavy rains to the left of the storm's center
track, while extreme winds are more common to the right of the track
(Halverson 2015).

Some storms can be of lower intensity (i.e., on the Saffir-Simpson
scale) yet bring dangerous rainfall, including well inland of the
storm's landfall. For example, storms including Floyd in 1999, Gaston in
2004, Irene in 2011, and Lee in 2011, have had severe inland impacts,
often through extreme rainfall, as post-tropical storms
{[}Halverson2015{]}. This rainfall can be particularly severe in the
Appalachians {[}Halverson2015{]}.

@@

@@

@@

\hypertarget{refs}{}
\hypertarget{ref-Halverson2015}{}
Halverson, Jeffrey B. 2015. ``Second Wind: The Dealdy and Destructive
Inland Phase of East Coast Hurricanes.'' \emph{Weatherwise} 68 (2):
20--27.
doi:\href{https://doi.org/10.1080/00431672.2015.997562}{10.1080/00431672.2015.997562}.

@@

>>>>

  \title{Assessing exposure to Atlantic Basin tropical storms in United States
counties}
    \author[Colorado State University]{Joshua Ferreri}
   \ead{joshua.m.ferreri@gmail.com}

@@

Current hurricane exposure assignments are often nonspecific and may
lead to missclassification of exposure at the county-level (Grabich et
al. 2016). Researchers have found heterogeneity in exposure assignments
of counties in assessing multiple classification methods, both within
and accross storms (i.e., storm dependent) (Grabich et al. 2016). Such
missclassification can have important public health and economic
implications.

Studies have begun to assess county-level hurricane exposure using large
historical datasets (Zandbergen 2009) \ldots{} making the
\texttt{hurricaneexposure} package a powerful tool in future exposure
analysis.

Currently, it seems that most hurricane track mapping capabilities are
restricted to geographical information system software. The introduction
of this package expands the mapping potential of hurricane tracks and
therefore the assessment of hurricane exposure and impacts to R.

@@

(Zandbergen 2009) developed an exposure factor based on location
(longitude and latitude), county shape, county size, and distance from
the coast. The study assumed symmetrical activity of storm related
factors,(Zandbergen 2009) which does not accurately represent storm
characteristics once landfall has been made.(Kruk et al. 2010, Halverson
(2015))

@@

Given its high resolution and consistancy with rain gage values,
(Villarini et al. 2011) recommends the use of Stage IV precipitation
data for the analysis of rainfall distribution in landfalling tropical
cyclones. However, the Stage IV dataset only provides data since 2001
(Lin and Mitchell 2005) and is therefore inadequate for use in this
analysis.

NLDAS does not provide information on precipitation over the ocean,
preventing its use in assessing the develpment of the storm prior to
landfall (Villarini et al. 2011). This point is moot for this analysis.

@@

Currently, distance parameters involved with assessing risk of a
particular storm have been rather arbitrary, contributing to the
necessity in understanding how such parameters influence a county's
exposure status. In assessing the risk of a given storm based on
distance, recent studies have defined the distance from the storm track
affected by a given storm differently. (Czajkowski, Simmons, and Sutter
2011) assessed county-level risk and exposure based on a three-tierd
definition, with primary counties being those closest to the storm track
on either side, secondary counties being adjacent to primary coutnies,
and tertiary counties adjacent to secondary counties. Such a defenition
resulted in an exposure defenition based on an average distance radius
of 120 km on either side of the storm track (Czajkowski, Simmons, and
Sutter 2011). Such a distance is slightly greater than that commonly
used by public health departments (i.e., 100 km) (Czajkowski, Simmons,
and Sutter 2011). In other cases, distance has been defined on the basis
of maximum sustained wind speeds and their decay, with exposure defined
as those counties exposed to wind speeds greater than or equal to
tropical storm strength (39--73 miles per hour) (Zandbergen 2009) and
may not provide an accurate respresentation of the communities exposed
to a given storm (e.g., a large number of inland counties are likely
underrepresented under this definition).

In a study assessing the association between hurricanes and undesireable
birth outcomes, researchers found that results were not sensitive to the
omission of residences 100 km from the storm path, and that results
varied insignificantly from 30-75 km (Currie and Rossin-Slater 2013).

@@

It can be very difficult to reliably measure rain during extreme rain
events, including tropical storms. For example, a heavy rain can wash
away {[}?{]} rain monitors {[}?{]}. It can also be very hard to measure
rain during heavy wind, as the rain does not fall straight into the
monitor {[}?{]}.

Some of the other possible sources for estimating rain during tropical
storms include \ldots{}{[}Stage IV, TMPA, NEXRAD{]}

@@

Rainfall estimates are likely underestimates for a few reasons. First,
they are based on averaging hourly measurements to a daily mean
estimate. This averaging would smooth over shorter periods of very
extreme rainfall. Further, this data is averaged over multiple grid
points within each county and so would not fully reflect very extreme
local precipitation (although this might be less of a concern for
classifying exposure to a large-scale storm system, like a tropical
storm, compared to more fine-scale storms). Finally, this NLDAS data
provides a re-analysis that incorporates measured rainfall, using
models, etc., to incorporate that observed data into a spatially and
temporally continuous dataset of rainfall. However, during extreme
storms, the problems with measuring rainfall using {[}rain monitors{]}
would propogate into the NLDAS data, so although NLDAS would prevent
missing values during the storm if monitors are not able to provide
data, if monitors are out, rainfall estimates from NLDAS will be based
more on models than on observations. More on NLDAS bias low here
(Villarini et al. 2011). Such bias, though unimportant in our analysis,
may play a larger role in the extrapolation to other studies or exposure
analysese where the precipitation cutoffs used here may provide a
threshold that results in the missclassification of less affected
counties as exposed.

Exposure classification based on rainfall has some advantages. For
example, it allows the identification of exposed counties that are
inland, rather than coastal, but that were affected by heavy rainfall
during the storm. Often, storm-related deaths are associated with inland
flooding. In fact, over half of the hurricane-related fatalities from
1970 to 1999 were a result of freshwater flooding, accounting for the
vast majority of deaths in inland counties ({\textbf{???}}). In another
analysis, researchers found that 79\% of freshwater-drowning fatalities
occured in inland counties,(Czajkowski, Simmons, and Sutter 2011)
further stressing the importance of rainfall in hurricane risk analysis.
Storms can produce a lot of rain especially in certain topographies,
like near mountains, so counties that are well inland can sometimes
experience more extreme rain that other counties at similar distances
from the storm's track.

In comparison to the storm track, the areas of extreme rain and extreme
wind can vary. For example, storms undergoing an extratropical
transition can bring heavy rains to the left of the storm's center
track, while extreme winds are more common to the right of the track
(Halverson 2015, Grabich et al. (2016)). In contrast, storm that
interact with a downstream ridge tend to bring heavy rains to the right
of the storm's center track (Atallah, Bosart, and Aiyyer 2007).

Some storms can be of lower intensity (i.e., on the Saffir-Simpson
scale) yet bring dangerous rainfall, including well inland of the
storm's landfall. For example, storms including Floyd in 1999, Gaston in
2004, Irene in 2011, and Lee in 2011, have had severe inland impacts,
often through extreme rainfall, as post-tropical storms (Halverson
2015). This rainfall can be particularly severe in the Appalachians
(Halverson 2015). Further, heavier rainfall is more common for storms
moving at slower forward speeds (Chang, Yang, and Kuo 2013). Storms with
a slower forward speed and larger rainfall area contribute to a longer
duration of rainfall, and therefore an increased chance of flood-related
damage (Rezapour and Baldock 2014).

@@

A recent study suggests that nearly all states east of the Rocky
Mountains have experienced wind exposure associated with either tropical
or post-tropical storms (Kruk et al. 2010).

@@

\paragraph{Table including various exposure papers and the parameters
used}\label{table-including-various-exposure-papers-and-the-parameters-used}

@@

\hypertarget{refs}{}
\hypertarget{ref-Atallah2007}{}
Atallah, Eyad, Lance F. Bosart, and Anantha R. Aiyyer. 2007.
``Precipitation Distribution Associated with Landfalling Tropical
Cyclones over the Eastern United States.'' \emph{Monthly Weather Review}
135: 2185--2206.
doi:\href{https://doi.org/10.1175/MWR3382.1}{10.1175/MWR3382.1}.

\hypertarget{ref-Chang2013}{}
Chang, Chih-Pei, Yi-Ting Yang, and Hung-Chi Kuo. 2013. ``Large
Increasing Trend of Tropical Cyclone Rainfall in Taiwan and the Roles of
Terrain.'' \emph{Journal of Climate} 26: 4136--47.
doi:\href{https://doi.org/10.1175/JCLI-D-12-00463.1}{10.1175/JCLI-D-12-00463.1}.

\hypertarget{ref-Currie2013}{}
Currie, Janet, and Maya Rossin-Slater. 2013. ``Weathering the Storm:
Hurricanes and Birth Outcomes.'' \emph{Journal of Health Economics} 32:
487--503.
doi:\href{https://doi.org/10.1016/j.jhealeco.2013.01.004}{10.1016/j.jhealeco.2013.01.004}.

\hypertarget{ref-Czajkowski2011}{}
Czajkowski, Jeffrey, Kevin Simmons, and Daniel Sutter. 2011. ``An
Analysis of Coastal and Inland Fatalities in Landfalling US
Hurricanes.'' \emph{Natural Hazards} 59: 1513--31.
doi:\href{https://doi.org/10.1007/s11069-011-9849-x}{10.1007/s11069-011-9849-x}.

\hypertarget{ref-Grabich2016}{}
Grabich, S. C., J. Horney, C. Konrad, and D. T. Lobdell. 2016.
``Measuring the Storm: Methods of Quantifying Hurricane Exposure with
Pregnancy Outcomes.'' \emph{Natural Hazards Review} 17 (1):
06015002--1--06015002--7.
doi:\href{https://doi.org/10.1061/(ASCE)NH.1527-6996.0000204}{10.1061/(ASCE)NH.1527-6996.0000204}.

\hypertarget{ref-Halverson2015}{}
Halverson, Jeffrey B. 2015. ``Second Wind: The Deadly and Destructive
Inland Phase of East Coast Hurricanes.'' \emph{Weatherwise} 68 (2):
20--27.
doi:\href{https://doi.org/10.1080/00431672.2015.997562}{10.1080/00431672.2015.997562}.

\hypertarget{ref-Kruk2010}{}
Kruk, Michael C., Ethan J. Gibney, David H. Levinson, and Michael
Squires. 2010. ``A Climatology of Inland Winds from Tropical Cyclones
for the Eastern United States.'' \emph{Journal of Applied Meteorology
and Climatology} 49: 1538--47.
doi:\href{https://doi.org/10.1175/2010JAMC2389.1}{10.1175/2010JAMC2389.1}.

@@

\hypertarget{ref-Lin2005}{}
Lin, Ying, and Kenneth E. Mitchell. 2005. ``The NCEP Stage II/IV Hourly
Precipitation Analyses: Development and Applications.'' \emph{19th
Conference of Hydrology, American Meteorlogical Society}.

\hypertarget{ref-Rezapour2014}{}
Rezapour, Mehdi, and Tom E. Baldock. 2014. ``Classification of Hurricane
Hazards: The Importance of Rainfall.'' \emph{Weather and Forcasting} 29:
1319--31.

\hypertarget{ref-Villarini2011}{}
Villarini, Gabriele, James A. Smith, Mary Lynn Baeck, Timothy Marchok,
and Gabriel A. Vecchi. 2011. ``Characterization of Rainfall Distribution
and Flooding Associated with U.S. Landfalling Tropical Cyclones:
Analysis of Hurricanes Frances, Ivan, and Jeanne (2004).'' \emph{Journal
of Geophysical Research} 116: D23116.

\hypertarget{ref-Zandbergen2009}{}
Zandbergen, Paul A. 2009. ``Exposure of US Counties to Atlantic Tropical
Storms and Hurricanes, 1851--2003.'' \emph{Natural Hazards} 48: 83--99.
doi:\href{https://doi.org/10.1007/s11069-008-9250-6}{10.1007/s11069-008-9250-6}.
